---
title: "Boundaries"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Boundaries}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(eval = FALSE)
```

A boundary says what a map is *about*: it draws the area's outline,
frames the view around it, and switches on the vignette that dims
everything else. Without one, the map simply fits the data:

```{r one-borough}
library(quickmap)                          # load the quickmap library
quickmap(
  "wandsworth_2017_2024_csv.csv",
  boroughs = "Wandsworth"                  # the boundary to draw
)
```

## 1. One, several, or all

```{r two-boroughs}
quickmap(
  from_rdata("bl_imperial_annualised_2021_2025_with_missing.Rdata", "no2",
             name = "Sensors"),
  boroughs = c("Merton", "Wandsworth")     # NEW: two boundaries
)
```

<iframe src="maps/boundaries-two.html" width="100%" height="440" loading="lazy" style="border:1px solid #ddd; border-radius:6px;" title="Two borough boundaries on one map"></iframe>

`boroughs = "All"` draws every London borough. Minor misspellings of
common names are corrected automatically; an unrecognised name lists the
valid ones.

## 2. Write the names on the map

On a multi-area map, label the areas themselves:

```{r boundary-labels}
quickmap(
  from_rdata("bl_imperial_annualised_2021_2025_with_missing.Rdata", "no2",
             name = "Sensors"),
  boroughs = c("Merton", "Wandsworth"),
  boundary_labels = TRUE                   # NEW: area names on the map
)
```

<iframe src="maps/boundaries-labelled.html" width="100%" height="440" loading="lazy" style="border:1px solid #ddd; border-radius:6px;" title="Borough names written on the map"></iframe>

## 3. The vignette

With a boundary drawn, everything outside it is dimmed by default, so
the reader's eye goes to the area in question. `vignette = FALSE` shows
the surroundings at full brightness; the dimming colour follows the
map's theme ([Styling and themes](styling.html)).

```{r vignette-off}
quickmap(
  "wandsworth_2017_2024_csv.csv",
  boroughs = "Wandsworth",
  vignette = FALSE                         # NEW: no dimming outside
)
```

## 4. No boundary at all

Leave `boroughs` out and no outline is drawn, no vignette applies, and
the view fits the data — right for data that isn't about one
administrative area:

```{r no-boundary}
quickmap("wandsworth_2017_2024_csv.csv")   # data-only map
```

## See also

- [Get started](quickmap.html) steps 3–4 — boundaries in the build-up.
- Ward-level boundaries and finer-grained labelling are on the roadmap
  (pre-v1.0 defect list) — this page will grow with them.
- `?quickmap` — `boroughs`, `boundary_labels`, `vignette`.
