---
title: "Labels"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Labels}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(eval = FALSE)
```

Labels put the number — or the site name — behind each symbol. One
parameter, `symbol_labels`, controls both what a label says and when it
appears:

```{r hover-values}
library(quickmap)                          # load the quickmap library
quickmap(
  "tubes_wandsworth.csv",
  boroughs      = "Wandsworth",
  symbol_labels = TRUE,                     # values appear on hover
  output_file = "wandsworth_labels.html"
)
```

<iframe src="maps/labels-hover.html" width="100%" height="440" loading="lazy" style="border:1px solid #ddd; border-radius:6px;" title="Hover a symbol to see its value"></iframe>

## 1. The five modes

| `symbol_labels =` | What appears | When |
|---|---|---|
| `FALSE` (default) | nothing | — |
| `TRUE` | the site's value | on hover |
| `"values_on"` | the site's value | always shown |
| `"labels"` | the site's name | on hover |
| `"labels_on"` | the site's name | always shown |

## 2. Always-visible values

For a map whose reader shouldn't have to hover — a report screenshot, a
public page — show the values permanently:

```{r values-on}
quickmap(
  "tubes_wandsworth.csv",
  boroughs      = "Wandsworth",
  symbol_labels = "values_on",              # NEW: values always shown
  output_file = "wandsworth_labels.html"
)
```

<iframe src="maps/labels-values-on.html" width="100%" height="440" loading="lazy" style="border:1px solid #ddd; border-radius:6px;" title="Values always shown beside each symbol"></iframe>

## 3. Site names instead of values

Where a label *says a name*, the name comes from the data — no extra
parameter, just the right column:

```{r names}
quickmap(
  list("tubes_wandsworth_labelled.csv", # tubes WITH a Label column
       "schools_wandsworth.csv"),                  # schools (School column)
  boroughs      = "Wandsworth",
  symbol_labels = "labels",                 # NEW: names on hover
  output_file = "wandsworth_labels.html"
)
```

<iframe src="maps/labels-names.html" width="100%" height="440" loading="lazy" style="border:1px solid #ddd; border-radius:6px;" title="Hover to see site and school names"></iframe>

## 4. How label content is chosen

Per layer, by its columns (duck typing):

| The layer has… | `"labels"` shows |
|---|---|
| a `School` column | the school's name |
| a `Label` column | the label text |
| neither | the site's value (with a warning) |

- **Gotcha:** in a CSV with a `Label` column, rows whose label cell is
  empty are dropped from the map entirely — fill or remove empty label
  cells.
- Mixed maps mix naturally: in the example above the tubes show their
  labels and the schools their names, each layer from its own column.

## 5. Shorter school names

School names repeat their type 53 times on a Merton map — "Abbotsbury
Primary School", "Aragon Primary School". The type is already in the
banner key, so `shorten_labels = TRUE` drops it from the markers:

```{r shorten}
quickmap(
  list("tubes_merton.csv",
       from_csv("schools_merton.csv",
                shorten_labels = TRUE)),      # NEW: "Abbotsbury", not
  boroughs      = "Merton",                   #      "Abbotsbury Primary School"
  symbol_labels = "labels",
  output_file   = "merton_schools.html"
)
```

It is set on the layer, not the map, because it describes that layer's
data. On Merton's 53 schools it shortens 47 and cuts the longest label
from 47 characters to 32.

**Two schools never merge into one label.** Where shortening would give
two schools the same name they keep their full names — "Harris Primary
Academy Merton" and "Harris Academy Merton" are different schools. The
exception is a pair whose `Level` differs: the cross colour already tells
them apart, so the short form is kept.

The vocabulary is schools-only (`Primary`, `Secondary`, `Infant`,
`Junior`, and a trailing `School`, `Academy` or `College`). Hospitals and
GP surgeries need their own, which is planned after 1.0.

## See also

- [Get started](quickmap.html) step 8 — labels in the build-up.
- [Your data](your-data.html) — where `Label` and `School` columns fit
  in each input format.
- `?quickmap` — the `symbol_labels` parameter.
