---
title: "Labels"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Labels}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(eval = FALSE)
```

Labels put the number — or the site name — behind each symbol. One
parameter, `marker_labels`, controls both what a label says and when it
appears:

```{r hover-values}
library(quickmap)                          # load the quickmap library
quickmap(
  "tubes_wandsworth.csv",
  boroughs      = "Wandsworth",
  marker_labels = TRUE                     # values appear on hover
)
```

<iframe src="https://ngnrfsk.github.io/quickmap/articles/maps/labels-hover.html" width="100%" height="440" loading="lazy" style="border:1px solid #ddd; border-radius:6px;" title="Hover a symbol to see its value"></iframe>

## 1. The five modes

| `marker_labels =` | What appears | When |
|---|---|---|
| `FALSE` (default) | nothing | — |
| `TRUE` | the site's value | on hover |
| `"values_on"` | the site's value | always shown |
| `"labels"` | the site's name | on hover |
| `"labels_on"` | the site's name | always shown |

## 2. Always-visible values

For a map whose reader shouldn't have to hover — a report screenshot, a
public page — show the values permanently:

```{r values-on}
quickmap(
  "tubes_wandsworth.csv",
  boroughs      = "Wandsworth",
  marker_labels = "values_on"              # NEW: values always shown
)
```

<iframe src="https://ngnrfsk.github.io/quickmap/articles/maps/labels-values-on.html" width="100%" height="440" loading="lazy" style="border:1px solid #ddd; border-radius:6px;" title="Values always shown beside each symbol"></iframe>

## 3. Site names instead of values

Where a label *says a name*, the name comes from the data — no extra
parameter, just the right column:

```{r names}
quickmap(
  list("tubes_wandsworth_labelled.csv", # tubes WITH a Label column
       "schools_wandsworth.csv"),                  # schools (School column)
  boroughs      = "Wandsworth",
  marker_labels = "labels"                 # NEW: names on hover
)
```

<iframe src="https://ngnrfsk.github.io/quickmap/articles/maps/labels-names.html" width="100%" height="440" loading="lazy" style="border:1px solid #ddd; border-radius:6px;" title="Hover to see site and school names"></iframe>

## 4. How label content is chosen

Per layer, by its columns (duck typing):

| The layer has… | `"labels"` shows |
|---|---|
| a `School` column | the school's name |
| a `Label` column | the label text |
| neither | the site's value (with a warning) |

- **Gotcha:** in a CSV with a `Label` column, rows whose label cell is
  empty are dropped from the map entirely — fill or remove empty label
  cells.
- Mixed maps mix naturally: in the example above the tubes show their
  labels and the schools their names, each layer from its own column.

## See also

- [Get started](quickmap.html) step 8 — labels in the build-up.
- [Your data](your-data.html) — where `Label` and `School` columns fit
  in each input format.
- `?quickmap` — the `marker_labels` parameter.
