---
title: "Labels"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Labels}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(eval = FALSE)
```

Labels put the number — or the site name — behind each symbol. One
parameter, `symbol_labels`, controls both what a label says and when it
appears:

```{r hover-values}
library(quickmap)                          # load the quickmap library
quickmap(
  "tubes_wandsworth.csv",
  boroughs      = "Wandsworth",
  symbol_labels = TRUE,                     # values appear on hover
  output_file = "wandsworth_labels.html"
)
```

<iframe src="https://ngnrfsk.github.io/quickmap/articles/maps/labels-hover.html" width="100%" height="440" loading="lazy" style="border:1px solid #ddd; border-radius:6px;" title="Hover a symbol to see its value"></iframe>

## 1. The five modes

| `symbol_labels =` | What appears | When |
|---|---|---|
| `FALSE` (default) | nothing | — |
| `TRUE` | the site's value | on hover |
| `"values_on"` | the site's value | always shown |
| `"labels"` | the site's name | on hover |
| `"labels_on"` | the site's name | always shown |

## 2. Always-visible values

For a map whose reader shouldn't have to hover — a report screenshot, a
public page — show the values permanently:

```{r values-on}
quickmap(
  "tubes_wandsworth.csv",
  boroughs      = "Wandsworth",
  symbol_labels = "values_on",              # NEW: values always shown
  output_file = "wandsworth_labels.html"
)
```

<iframe src="https://ngnrfsk.github.io/quickmap/articles/maps/labels-values-on.html" width="100%" height="440" loading="lazy" style="border:1px solid #ddd; border-radius:6px;" title="Values always shown beside each symbol"></iframe>

## 3. Site names instead of values

Where a label *says a name*, the name comes from the data — no extra
parameter, just the right column:

```{r names}
quickmap(
  list("tubes_wandsworth_labelled.csv", # tubes WITH a Label column
       "schools_wandsworth.csv"),                  # schools (School column)
  boroughs      = "Wandsworth",
  symbol_labels = "labels",                 # NEW: names on hover
  output_file = "wandsworth_labels.html"
)
```

<iframe src="https://ngnrfsk.github.io/quickmap/articles/maps/labels-names.html" width="100%" height="440" loading="lazy" style="border:1px solid #ddd; border-radius:6px;" title="Hover to see site and school names"></iframe>

## 4. How label content is chosen

Per layer, by its columns (duck typing):

| The layer has… | `"labels"` shows |
|---|---|
| a `School` column | the school's name |
| a `Label` column | the label text |
| neither | the site's value (with a warning) |

- **Gotcha:** in a CSV with a `Label` column, rows whose label cell is
  empty are dropped from the map entirely — fill or remove empty label
  cells.
- Mixed maps mix naturally: in the example above the tubes show their
  labels and the schools their names, each layer from its own column.

Because the choice is made per layer, one setting can do two jobs.
`symbol_labels = "values_on"` on a map of tubes *and* schools gives every
tube its value and every school its name — even if the tube file also has
a `Label` column, since a `School` column always wins. Values read
`28 µg/m³`.

## 5. Shorter school names

School names repeat their type 53 times on a Merton map — "Abbotsbury
Primary School", "Aragon Primary School". The type is already in the
banner key, so `shorten_labels = TRUE` drops it from the markers:

```{r shorten}
quickmap(
  list("tubes_merton.csv",
       from_csv("schools_merton.csv",
                shorten_labels = TRUE)),      # NEW: "Abbotsbury", not
  boroughs      = "Merton",                   #      "Abbotsbury Primary School"
  symbol_labels = "labels",
  output_file   = "merton_schools.html"
)
```

It is set on the layer, not the map, because it describes that layer's
data. On Merton's 53 schools it shortens 47 and cuts the longest label
from 47 characters to 32.

**Two schools never merge into one label.** Where shortening would give
two schools the same name they keep their full names — "Harris Primary
Academy Merton" and "Harris Academy Merton" are different schools. The
exception is a pair whose `Level` differs: the cross colour already tells
them apart, so the short form is kept.

The vocabulary is schools-only (`Primary`, `Secondary`, `Infant`,
`Junior`, and a trailing `School`, `Academy` or `College`). Hospitals and
GP surgeries need their own, which is planned after 1.0.

## 6. Size and background

Labels are the smallest text on a map, matched to the smallest text in
the legend, and they scale with a static export automatically — a
4000-pixel image gets 4000-pixel-sized labels without being asked.

Two theme keys adjust them:

```yaml
map:
  label_scale: 1           # multiplier: 2 is twice the legend's small text
  label_background: true   # the translucent plate behind each label
```

`label_background: false` removes the plate, leaving the text alone on
the map. The plate buys legibility over busy tiles and costs clutter
where labels are dense — worth turning off on a print map.

`label_scale` is a trade, not a free win: bigger labels collide sooner.
On a crowded borough map, labels twice this size begin to overlap in the
built-up centre, and much beyond that they cover the map. If you need
larger text on paper, print the image bigger before raising this — the
same file across A3 is twice the point size for nothing.

## 7. Reference layers get a key, not labels

A layer with no values — schools are the usual case — has no place on
the colour ramp, so its symbol is explained in the banner instead. Where
such a layer has a `Level` column, its categories appear at the end of
the banner as small marks in their own colours ("✕ Primary ✕ Secondary").

That key is what lets the map labels drop the words every school shares:
it is why `shorten_labels = TRUE` above is safe.

## See also

- [Get started](quickmap.html) step 8 — labels in the build-up.
- [Your data](your-data.html) — where `Label` and `School` columns fit
  in each input format.
- `?quickmap` — the `symbol_labels` parameter.
- [Sharing and export](sharing.html) — sizing labels for the printed
  page.
