---
title: "Layers"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Layers}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(eval = FALSE)
```

A QuickMap map is a stack of **layers**, one per data source. Measurement
layers (tubes, sensors) draw as solid coloured symbols and drive the time
animation; context layers (schools, hospitals, proposed sites) draw as
outline symbols on top. The stack is the first argument of `quickmap()` —
a single file, or a list:

```{r two-lines}
library(quickmap)                          # load the quickmap library
quickmap(
  list("tubes_merton.csv",          # diffusion tube data
       "schools_merton.csv"),              # school locations
  boroughs = "Merton",                      # the borough boundary to draw
  output_file = "merton_layers.html"
)
```

The tubes CSV needs `Easting`, `Northing` and one column per year of
readings; the schools CSV needs `Easting`, `Northing` and a `School`
column. QuickMap recognises what each file contains from its columns —
you did not have to say which was which. (Full input specifications:
the *Your data* chapter.)

Each numbered step below adds one idea to the previous one.

## 1. Use the `layers` list to add sources

Comparing monitoring networks, or adding context around them, means
putting more sources on one map — just extend the list:

```{r three-layers}
quickmap(
  list(
    "tubes_merton.csv",                            # tubes -> circles
    "sensors_london.RData", # sensors -> diamonds
    "schools_merton.csv"                                   # schools -> crosses
  ),
  boroughs = "Merton",                     # the borough boundary to draw
  title = "Merton NO2: ● tubes ◆ sensors ✖ schools", # symbols key the banner
  output_file = "merton_layers.html"
)
```

The new input: an RData sensor export, needing `siteCode`, `lat`, `lon`,
a year column and a pollutant column (`no2` here). Each layer keeps its
own symbol, so readers can tell a reference tube from a low-cost sensor
at a glance. The time slider covers every year found in *any* layer;
layers without data for a year simply sit that year out.

<iframe src="https://ngnrfsk.github.io/quickmap/articles/maps/layers-multilayer.html" width="100%" height="440" loading="lazy" style="border:1px solid #ddd; border-radius:6px;" title="Three layers on one map: circles (tubes), diamonds (sensors), crosses (schools)"></iframe>

## 2. Name a layer with `from_csv(name = )`

The legend and controls name layers after their filenames. To choose the
name yourself, build the layer with `from_csv()` instead of a bare path:

```{r named-layer}
quickmap(
  list(
    from_csv("tubes_merton.csv",
             name = "Council tubes"),      # your name, not the filename
    "schools_merton.csv"                   # untouched layers stay as paths
  ),
  boroughs = "Merton",
  output_file = "merton_layers.html"
)
```

This is the general pattern for customising one layer of several: **wrap
that layer, leave the rest alone**.

## 3. Choose each layer's pollutant with `from_rdata()`

Sensor files usually carry several pollutants; the second argument of
`from_rdata()` selects one:

```{r rdata-pollutant}
quickmap(
  from_rdata("sensors_london.RData",
             "pm25",                       # the pollutant column to map
             name = "Breathe London"),
  boroughs = "Merton",
  colour_scale = "gla_pm25",                # a PM2.5 colour scale
  output_file = "merton_layers.html"
)
```

## 4. Fetch data live with `from_openair()`

No file at all: the [openair package](https://openair-project.github.io/openair/)
downloads UK monitoring data, and `from_openair()` turns its output
straight into a layer:

```{r net-openair-live}
library(openair)                           # the OpenAir data tools
kc1 <- importUKAQ(site = "kc1",            # North Kensington AURN site
                  year = 2023)             # one year of hourly data
quickmap(
  from_openair(kc1,
               pollutant = "no2",          # pollutant column to map
               avg.time = "month",         # aggregate hours -> monthly means
               source = "aurn",            # network, for coordinate lookup
               name = "AURN"),             # layer name for the map
  boroughs = "All",                         # all London borough boundaries
  output_file = "merton_layers.html"
)
```

This example carries a second, smaller idea: `avg.time = "month"`
aggregates the hourly download to monthly means before mapping —
the vocabulary ("year", "month", "day", "hour") is openair's own. See
the [OpenAir book](https://openair-project.github.io/book/) for
everything `importUKAQ()` can fetch.

## 5. Map your own data frame with `qm_layer()`

Any data frame with a site code, coordinates and a value column can be a
layer — this is how you map data QuickMap has never heard of:

```{r hand-built}
my_sites <- data.frame(
  code = c("A", "B", "C"),                 # a stable id per site (required)
  lat  = c(51.44, 51.45, 51.46),           # latitude, WGS84 (required)
  lon  = c(-0.17, -0.19, -0.21),           # longitude, WGS84 (required)
  no2  = c(18, 27, 41)                     # the values to map (required)
)
quickmap(
  qm_layer(my_sites, name = "My survey"),  # wrap the data frame as a layer
  boroughs = "Wandsworth",
  output_file = "merton_layers.html"
)
```

## 6. Choose each layer's symbol with `qm_layer(shape = )`

When two layers would otherwise look alike, give one a different symbol:

```{r choose-shape}
quickmap(
  list(
    from_csv("tubes_merton.csv",
             name = "Council tubes"),      # circles (the tube default)
    qm_layer(my_sites,
             shape = "star",               # this layer draws as stars
             name = "My survey")
  ),
  boroughs = "Merton",
  output_file = "merton_layers.html"
)
```

Friendly names work: `"circle"`, `"square"`, `"diamond"`, `"triangle"`,
`"cross"`, `"star"`, `"plus"` — a misspelling gives an error listing every
valid name.

## 7. How it works

Two rules carry the whole system:

1. **Layer properties live on the layer; map properties live on
   `quickmap()`.** A layer's value column, time column, labels, symbol
   shape and name are set by `qm_layer()` or a `from_*()` wrapper. The
   map's boroughs, colour scale, title, theme and output file are
   `quickmap()` arguments. (One historic exception survives: the
   map-level `data_symbols` vector, kept as an override.)
2. **Content decides, not filenames (duck typing).** A CSV with a `School`
   column is a school layer; a CSV with several year columns (`2017`,
   `2018`, …) is a temporal tube layer; an RData file is searched for a
   compatible sensor table. Rename your files freely.

## 8. Full detail

| Function | Builds a layer from | Key arguments |
|---|---|---|
| `from_csv()` | CSV file (tubes or schools) | `pollutant`, `name`, `temporal` |
| `from_rdata()` | RData sensor export | `pollutant`, `name`, `data_object_name` |
| `from_openair()` | openair tibble | `pollutant`, `avg.time`, `source`, `name` |
| `qm_layer()` | any data frame / sf object | `value_col`, `time_col`, `label_col`, `shape`, `name` |
| `qm_meta()` | — | inspect a layer's metadata |

See each function's help page (e.g. `?qm_layer`) for the complete
contract. Cross-cutting behaviour that no single help page owns:

- **Required identity column:** every layer needs a stable site identifier —
  `code` (or legacy `siteCode`) for data frames; CSVs get one generated
  from coordinates or the `Label` column.
- **Column aliases:** `siteCode`→`code`, `year_str`→`time_label`,
  `Latitude`/`Longitude`→`lat`/`lon` are accepted and normalised.
- **Shapes** (v0.9.8.1+): set per layer with `qm_layer(shape = )`, or
  inherited from the `from_*()` conventions (tubes circle, sensors
  diamond, schools cross). Layers with no shape get an automatic
  assignment — solid symbols for time-varying layers, outline symbols
  for static ones, cycling in layer order. A map-level
  `quickmap(data_symbols = c(...))` vector (one shape per layer)
  overrides everything.
- **Gotcha — the `Label` column:** in a CSV with a `Label` column, rows
  whose label cell is empty are dropped from the map entirely.
- **Coordinates:** CSVs use British National Grid (`Easting`/`Northing`);
  data frames use WGS84 `lat`/`lon`; sf objects keep their CRS.

## See also

- [Get started](quickmap.html) — the first map, step by step, up to a
  full animation.
- `?quickmap` — every map-level parameter.
- [OpenAir book](https://openair-project.github.io/book/) and
  [worldmet](https://openair-project.github.io/worldmet/) — fetching
  measurement and met data (worked extraction examples arrive with the
  *Your data* chapter).
