---
title: "Get started"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Get started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(eval = FALSE)
```

QuickMap makes air quality maps from the data you already have — a
diffusion tube file, a sensor-network export, or a live data source such
as an [OpenAir](https://openair-project.github.io/book/) data pull. Maps
are self-contained files, so you can email them or put them on a website,
and compact, fast animations are easy to create.

This page builds one step at a time, from the most basic map to an
animated pollution episode with a live wind overlay. The maps shown are
real QuickMap output — click, drag and play with them.

## 1. Install

Install from GitHub:

```{r install, purl = FALSE}
# install.packages("devtools")            # if you don't have it
devtools::install_github("ngnrfsk/quickmap")
```

> While the repository is private, `install_github()` needs a GitHub
> personal access token: create one at github.com → Settings → Developer
> settings, then run `Sys.setenv(GITHUB_PAT = "your-token")` first. This
> step disappears when the package is released.

## 2. Make your first map — one argument

```{r first-map, purl = FALSE}
library(quickmap)                           # load the quickmap library
quickmap(
  "~/my_data/wandsworth_2017_2024_csv.csv"  # diffusion tube data
)
```

One input — the measurement data — is all `quickmap()` needs. The data
here is an ordinary diffusion tube spreadsheet saved as CSV:

| Easting | Northing | 2017 | 2018 | ... | 2024 | Label |
|---------|----------|------|------|-----|------|-------|
| 528900  | 172431   | 33   | 31   | ... | 15.9 | 11b Elmbourne Road |
| 530129  | 177727   |      | 49   | ... | 22.9 | 1 Nine Elms |

*Minimum required: `Easting`, `Northing` and at least one year column.
Anything else is optional. (Full input specifications: the* Your data
*chapter.)*

QuickMap converts the coordinates, colours each site by the WHO NO2
bands, fits the view to the data, and writes `aq_maps/quickmap.html` —
a single self-contained file. Open it in any browser, attach it to an
email, or put it on a website (it does need an internet connection to
download the background map). Here is that exact file:

<iframe src="maps/getstarted-first.html" width="100%" height="440" loading="lazy" style="border:1px solid #ddd; border-radius:6px;" title="First map: Wandsworth diffusion tubes, no boundary"></iframe>

## 3. Draw a borough boundary

Air quality reporting is usually about a place; naming it draws its
boundary and frames the map around it:

```{r boundary, purl = FALSE}
quickmap(
  "~/my_data/wandsworth_2017_2024_csv.csv",
  boroughs = "Wandsworth"                   # the boundary to draw
)
```

`boroughs` takes one name, several (`c("Wandsworth", "Richmond")`), or
`"All"` for every London borough.

<iframe src="maps/getstarted-boundary.html" width="100%" height="440" loading="lazy" style="border:1px solid #ddd; border-radius:6px;" title="The same map with the Wandsworth boundary and vignette"></iframe>

## 4. The vignette — focus attention inside the boundary

Notice the map above dims everything outside Wandsworth. That is the
**vignette**, on by default whenever a boundary is drawn, so the reader's
eye goes to the area the map is about. To show the surroundings at full
brightness instead:

```{r vignette, purl = FALSE}
quickmap(
  "~/my_data/wandsworth_2017_2024_csv.csv",
  boroughs = "Wandsworth",
  vignette = FALSE                          # no dimming outside the boundary
)
```

## 5. Save typing: tell QuickMap where your data lives

Optional, but worth doing once: if your data files sit in one folder, set
`DATA_PATH` and use bare filenames everywhere (the rest of this page
assumes it is set):

```{r datapath, purl = FALSE}
Sys.setenv(
  DATA_PATH = "~/my_data"                  # folder where your data lives
)
quickmap(
  "wandsworth_2017_2024_csv.csv",          # now just the filename
  boroughs = "Wandsworth"
)
```

## 6. Reading the map

- **Symbols** are monitoring sites, coloured by pollution level — green
  is good, red is bad. Hover for details.
- **The legend** (along the bottom) names each colour band. Click its
  header to collapse it.
- **The time slider** (bottom of the map) appears because your CSV had
  several year columns. Drag it, step with ‹ ›, or press play to animate.
- **The banner** (top) is the map's title — added next.

## 7. Add a title and choose the file name

Maps usually need a title for the banner, and a file name you'll
recognise later:

```{r step-further}
library(quickmap)                          # (already loaded if you followed above)
quickmap(
  "wandsworth_2017_2024_csv.csv",          # same data
  boroughs    = "Wandsworth",
  title       = "Wandsworth NO2, 2017-2024", # banner text
  output_file = "wandsworth_no2.html"        # file written to aq_maps/
)
```

## 8. Combine several measurement networks

Different monitoring networks — and context like schools — go on one map
by listing the files. Each network keeps its own symbol (tubes circles,
sensors diamonds, schools crosses), so readers can tell them apart:

```{r networks}
quickmap(
  list(
    "merton_dt_2018_2024.csv",                             # diffusion tubes
    "bl_imperial_annualised_2021_2025_with_missing.Rdata", # sensor network
    "schools_Merton.csv"                                   # school locations
  ),
  boroughs = "Merton",
  title = "Merton NO2: tubes, sensors and schools"
)
```

<iframe src="maps/getstarted-networks.html" width="100%" height="440" loading="lazy" style="border:1px solid #ddd; border-radius:6px;" title="Three networks on one map"></iframe>

Naming layers, choosing each network's pollutant and symbol, and mapping
your own data frames: the [Layers](layers.html) chapter.

## 9. Make a sophisticated animation

The same call scales from eight annual frames to a full pollution
episode. This maps five days of hourly sensor data — 108 time steps —
and starts playing by itself. Press play, or drag the slider:

```{r animation}
quickmap(
  from_rdata("episodeJan15-20_2024_sf_all.Rdata", # hourly sensor export
             "pm25",                              # pollutant to map
             name = "Sensors"),                   # layer name for the map
  boroughs     = c("Wandsworth", "Richmond"),     # two boundaries
  colour_scale = "stripes_pm25",                  # a PM2.5 colour scale
  title        = "PM2.5 Episode: Jan 15-20, 2024",
  output_file  = "episode.html",
  autoplay     = TRUE,                            # play on open
  play_speed   = 500                              # milliseconds per hour
)
```

<iframe src="maps/getstarted-episode.html" width="100%" height="480" loading="lazy" style="border:1px solid #ddd; border-radius:6px;" title="108-step hourly animation"></iframe>

Above 50 time steps QuickMap automatically switches to a compact storage
format, so even this five-day hourly animation stays around 1 MB — small
enough to email.

## 10. Add live wind to the animation

Where did the pollution come from? An animated wind-particle overlay
answers that at a glance. `from_worldmet()` fetches real weather-station
data (here Heathrow) from NOAA, and the particles advance with the time
slider:

```{r net-wind}
heathrow <- from_worldmet(station = "037720-99999", # Heathrow station id
                          year = 2024)              # fetches from NOAA
quickmap(
  from_rdata("episodeJan15-20_2024_sf_all.Rdata", "pm25", name = "Sensors"),
  boroughs     = c("Wandsworth", "Richmond"),
  colour_scale = "stripes_pm25",
  title        = "PM2.5 Episode with wind: Jan 15-20, 2024",
  output_file  = "episode_wind.html",
  autoplay     = TRUE,
  play_speed   = 500,
  wind         = heathrow                           # the wind overlay
)
```

<iframe src="maps/getstarted-wind.html" width="100%" height="480" loading="lazy" style="border:1px solid #ddd; border-radius:6px;" title="The animation with a live wind-particle overlay"></iframe>

Wind particles are coloured by wind speed. (Requires the `worldmet`
package and an internet connection for the NOAA fetch; more in the
*Wind* chapter.)

## More things QuickMap can do

One line each; every one has a chapter or help page:

- **Colour scales** — WHO bands, GLA PM2.5, borough-specific scales:
  `colour_scale = "gla_pm25"` (`?load_colour_scale`).
- **Themes** — banner style (white strip or colour bar), base map tiles
  (pale or OpenStreetMap), colours, autoplay — one reusable YAML file:
  `theme_file = "my_theme.yaml"` (`?load_theme`).
- **Symbol labels** — values or site names, on hover or always:
  `marker_labels = "labels"`.
- **Static JPG export** for reports: `export_image = c(1600, 1200)` —
  text and legend scale with the image size.
- **Choose the time steps**: `display_times = 2020:2022`.
- **Map your own data frame** with `qm_layer()` — any data with a site
  code, coordinates and values ([Layers](layers.html)).

## Where next

- **[Layers](layers.html)** — networks, naming, symbols, your own data.
- **Function help** — `?quickmap` lists every parameter.
- **Fetching data with OpenAir** — no data file yet? The
  [OpenAir project](https://openair-project.github.io/book/) can download
  UK monitoring data for you, and QuickMap maps it directly (worked
  examples coming in the *Your data* chapter).
