---
title: "For R users: the layer contract"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{For R users: the layer contract}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(eval = FALSE)
```

Under every input format sits one internal currency: the `qm_layer` — an
`sf` object in long format (one row per site per time step) carrying its
metadata as attributes. The `from_*()` wrappers all produce one; this
page is the contract for building and inspecting them directly.

## 1. Canonical columns

| Column | Type | Meaning |
|---|---|---|
| `code` | character | stable site id, links a symbol across time steps |
| `time_label` | character | display text per step (absent = static layer) |
| `time_sort` | POSIXct | sort key parsed from the label |
| *value column* | numeric/character | the mapped values, named by `value_col` |
| `lat`, `lon` | numeric | WGS84 |
| `geometry` | sfc | the sf geometry (built from lat/lon if absent) |

Legacy names are accepted and normalised: `siteCode`→`code`,
`year_str`→`time_label`, `Latitude`/`Longitude`→`lat`/`lon`.

## 2. Construction and inspection

```{r construct}
library(quickmap)
d <- data.frame(
  code = rep(c("A", "B"), each = 3),
  date = rep(as.Date(c("2024-01-01", "2024-02-01", "2024-03-01")), 2),
  no2  = c(31, 28, 25, 44, 41, 37),
  lat  = rep(c(51.44, 51.46), each = 3),
  lon  = rep(c(-0.17, -0.21), each = 3)
)
layer <- qm_layer(d, name = "survey")      # infers value, time, resolution
qm_meta(layer)                             # value_col, time_col, shape, ...
print(layer)                               # "2 sites x 3 time steps [month]"
```

Explicit overrides: `value_col`, `time_col` (pass it for non-date
orderings like diurnal hours), `label_col`, `shape`, `name`.

## 3. The time grammar

`time_label` values must all share one format; the format sets the
animation's resolution:

| Format | Resolution |
|---|---|
| `YYYY` | year |
| `YYYY-MM` | month |
| `YYYY-MM-DD` | day |
| `YYYY-MM-DD HH:MM(:SS)` | hour/minute |

`Date`/`POSIXct` columns are recognised by class wherever they appear;
character columns are only parsed if named `date`, `time`, `datetime`,
`time_label` or `year_str` — a gate-named column that fails the grammar
is a loud error, never a silent guess.

## 4. Inference, in order

When not told explicitly, `qm_layer()` infers:

- **value column**: a known pollutant name (`no2`, `pm25`, `pm10`, `nox`,
  `o3`, …), else the single remaining candidate column — two candidates
  is an error asking for `value_col`.
- **time column**: by class, then by the name gate above; none = a
  static layer.
- **labels**: a `Label` or `School` column if present.
- **shape**: none — the map assigns symbols (solid for temporal layers,
  outline for static, in layer order) unless the layer or the map-level
  `data_symbols` says otherwise.

## 5. The compatibility wrapper

`create_pollution_map()` is the historic entry point, kept working with
its original signature; it converts `data_sources` to `qm_layer`s and
delegates to `quickmap()`. New code should call `quickmap()`; old scripts
need no changes.

## 6. Why not tmap, mapview or openairmaps?

A fair question for anyone already fluent in R spatial packages. QuickMap
does not replace them; it fills one gap they leave.

**tmap** gives a full grammar over `sf` and is highly customisable for
cartography. Two things rule it out here. Its animation route
(`tm_facets()` + `tmap_animation()`) renders GIF or video, not
self-contained interactive HTML with a time control, so the grammar would
not buy the temporal feature that is the point of this package. And the
entry cost is high for the intended reader: shape/layer separation, `+`
composition and the visual-variable system are three concepts before the
first map appears.

**mapview** has the gentlest curve in the R spatial ecosystem —
`mapview(x, zcol = "no2")` renders immediately — and QuickMap borrows its
central lesson: keep the atomic unit a standard `sf` object the user can
build and inspect, rather than inventing a class. But mapview has no
temporal animation, and it is built for exploring data rather than
producing publication output.

**openairmaps** is the nearest neighbour in subject matter: leaflet maps
of UK air quality networks, polar-analysis markers, trajectory paths, with
static equivalents for reports. It maps analysis markers and network
locations; it does not produce animated concentration maps with
configurable banner, legend and controls.

**Symbol control** was the third consideration. These maps routinely carry
circles for diffusion tubes, diamonds for sensors and crosses for schools
in one figure, on one colour scale, with the symbol set per layer. None of
the packages above exposes that control in a form that survives into both
interactive and static output.

What is left is **leaflet**, which QuickMap builds on directly. Everything
above the base map — banner, colour-ramp legend, time slider, network
indicator, attribution — is QuickMap's own, injected into the saved HTML.

The API survey is in the repository, at
`dev/260706_atomic_unit_recommendation.md`. The rendering backends
(Leaflet, MapLibre/mapgl, mapdeck) were benchmarked separately in July
2026, with MapLibre recorded as the migration path beyond v1.

## See also

- [Layers](layers.html) — the user-facing view of the same system.
- `?qm_layer`, `?qm_meta` — the reference.
- [Your data](your-data.html) — the file formats the wrappers consume.
