---
title: "Recipes"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Recipes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(eval = FALSE)
```

Complete, copy-paste starting points for common jobs. Each recipe is one
call; swap in your files and names.

## 1. The annual borough report map

Tubes + sensors + schools, borough theme, labels on hover, JPG for the
report — the standard local-authority deliverable:

```{r recipe-report}
library(quickmap)                          # load the quickmap library
quickmap(
  list(
    "tubes_merton.csv",                             # diffusion tubes
    "sensors_london.RData", # sensor network
    "schools_merton.csv"                                   # schools
  ),
  boroughs      = "Merton",
  title         = "Merton NO2: ● tubes ◆ sensors ✖ schools",
  theme_file    = system.file("themes", "merton.yaml",
                              package = "quickmap"),
  symbol_labels = "labels",                # names on hover, where the data
                                           # has them (layers without a
                                           # Label column show values)
  display_times = 2020:2022,
  export_image  = TRUE,                    # a JPG per year, 1200x1200
  output_file   = "merton_report.html"
)
```

## 2. The pollution-episode animation

Hourly sensor data across two boroughs, autoplaying, with wind:

```{r recipe-episode, purl = FALSE}
heathrow <- from_worldmet(station = "037720-99999", year = 2024)
quickmap(
  from_rdata("episode_london.RData", "pm25", name = "Sensors"),
  boroughs     = c("Wandsworth", "Richmond"),
  colour_scale = "stripes_pm25",
  title        = "PM2.5 Episode: Jan 15-20, 2024",
  autoplay     = TRUE,
  play_speed   = 500,
  wind         = heathrow,
  output_file  = "episode.html"
)
```

## 3. Proposed monitoring sites over existing ones

Site-selection maps: current networks plus a static CSV of proposed
locations (any CSV with `Easting`/`Northing` and no year columns is a
context layer):

```{r recipe-proposed}
quickmap(
  list(
    "tubes_wandsworth.csv",        # existing tubes
    "sensors_london.RData", # existing sensors
    "proposed_sites_wandsworth.csv"      # proposed sites (static CSV;
                                           # its School column supplies names)
  ),
  boroughs      = "Wandsworth",
  title         = "Wandsworth sites: ● tubes ◆ sensors ✖ proposed",
  symbol_labels = "labels",
  output_file   = "wandsworth_sites.html"
)
```

## 4. Straight from the internet: fetch-to-map with OpenAir

No local data at all — download a year of reference-monitor data and map
it monthly, in one pipeline:

```{r net-recipe-fetch}
library(openair)                           # the OpenAir data tools
quickmap(
  from_openair(importUKAQ(site = "kc1", year = 2023), # download AURN data
               pollutant = "no2",
               avg.time = "month",         # monthly means
               source = "aurn",
               name = "AURN"),
  boroughs    = "All",
  title       = "North Kensington NO2, monthly 2023",
  output_file = "kc1_monthly.html"
)
```

## See also

- [Get started](quickmap.html) — what each parameter here does, step by
  step.
- [Your data](your-data.html) §5 — more OpenAir extraction patterns.
