---
title: "Styling and themes"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Styling and themes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(eval = FALSE)
```

Two systems style a QuickMap map: a **colour scale** decides what the
data's colours mean, and a **theme** decides how everything else looks —
banner, background tiles, vignette, controls. Both are plain YAML files
you can copy and edit.

```{r scale}
library(quickmap)                          # load the quickmap library
quickmap(
  from_rdata("sensors_london.RData",
             "pm25",                       # a PM2.5 map...
             name = "Sensors"),
  boroughs     = "Merton",
  colour_scale = "gla_pm25"                # ...needs a PM2.5 scale
)
```

<iframe src="https://ngnrfsk.github.io/quickmap/articles/maps/styling-scale.html" width="100%" height="440" loading="lazy" style="border:1px solid #ddd; border-radius:6px;" title="PM2.5 mapped with the GLA scale"></iframe>

## 1. The bundled colour scales

| Name | Pollutant | Bands |
|------|-----------|-------|
| `who_no2` (default) | NO2 | WHO guideline bands |
| `stripes_no2` | NO2 | stripe-style |
| `stripes_pm25` | PM2.5 | stripe-style |
| `airstat_no2` | NO2 | AirStat |
| `gla_pm25` | PM2.5 | GLA |
| `lbw_no2`, `lbrut_no2`, `lbm_no2` | NO2 | borough-specific (Wandsworth, Richmond, Merton) |
| `deltas` | any | year-on-year change |
| `schools` | — | categorical (school type) |

`load_colour_scale("who_no2")` shows any scale's definition. (Two further
project-specific scales, `nine_elms` and `nine_elms_consolidated`, also
ship with the package.)

## 2. Anatomy of a scale — make your own

A scale is one YAML file: thresholds, colours, labels. Copy a bundled one
(`inst/config/scales/` in the package) and edit:

```yaml
name: my_no2
title: "NO2, µg/m³"
pollutant: NO2
thresholds: [0, 10, 20, 40, .Inf]
colours: ["green", "yellow", "orange", "red", "white"]
labels: ["< 10: Good", "10-20: Fair", "20-40: Poor", "> 40: Bad", "No data"]
```

The legend is generated from this file — labels with a description after
the range (like "10: Good") gain a footnote symbol and appear in the
legend's key row.

## 3. Themes — one file for the whole look

A theme bundles every visual choice so a series of maps stays consistent:

```{r theme}
quickmap(
  "tubes_merton.csv",
  boroughs   = "Merton",
  theme_file = system.file("themes", "merton.yaml", # NEW: a bundled theme
                           package = "quickmap")
)
```

<iframe src="https://ngnrfsk.github.io/quickmap/articles/maps/styling-theme.html" width="100%" height="440" loading="lazy" style="border:1px solid #ddd; border-radius:6px;" title="The Merton borough theme"></iframe>

The main sections (all optional — anything omitted keeps its default;
there is also a `legend:` section and a free-form `palette:` of named
colours):

```yaml
banner:
  style: "strip"          # "strip" (white, brand rule) or "bar" (colour block)
  background: "#5F3E94"   # the brand colour
  title: "Merton Air Quality"
map:
  vignette: true
  base_tiles: null        # null = OpenStreetMap; or e.g. "CartoDB.Positron"
  boundary_labels: false
controls:
  autoplay: false
  play_speed: 500
wind:                     # wind-particle styling (see the Wind chapter)
  particle_density: 0.0033
  line_width: 1.5
```

Bundled themes: `merton`, `wandsworth`, `richmond`, `airstat`,
`high_contrast` (`show_borough_colours()` lists them with their
palettes).

## 4. Who wins: parameter, theme, or default

Explicit `quickmap()` arguments beat the theme; the theme beats the
built-in defaults. So a theme sets the house style, and any single map
can override one thing:

```{r precedence, purl = FALSE}
quickmap(
  "tubes_merton.csv",
  boroughs      = "Merton",
  theme_file    = "my_theme.yaml",         # house style...
  banner_colour = "#078141"                # ...this map's banner differs
)
```

## 5. The background tiles

`map: base_tiles:` in the theme takes any
[Leaflet provider name](https://leaflet-extras.github.io/leaflet-providers/preview/).
The default (OpenStreetMap) carries the vignette well; the pale
`"CartoDB.Positron"` suits dense data.

## See also

- [Get started](quickmap.html) steps 4 and 7 — vignette colour and
  `banner_colour` in the build-up.
- [Wind](wind.html) — the theme's `wind:` section.
- `?load_theme`, `?load_colour_scale`, `?show_borough_colours`.
