---
title: "Styling and themes"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Styling and themes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(eval = FALSE)
```

Two systems style a QuickMap map: a **colour scale** decides what the
data's colours mean, and a **theme** decides how everything else looks —
banner, background tiles, vignette, controls. Both are plain YAML files
you can copy and edit.

```{r scale}
library(quickmap)                          # load the quickmap library
quickmap(
  from_rdata("sensors_london.RData",
             "pm25",                       # a PM2.5 map...
             name = "Sensors"),
  boroughs     = "Merton",
  colour_scale = "gla_pm25",                # ...needs a PM2.5 scale
  output_file = "merton_styled.html"
)
```

<iframe src="https://ngnrfsk.github.io/quickmap/articles/maps/styling-scale.html" width="100%" height="440" loading="lazy" style="border:1px solid #ddd; border-radius:6px;" title="PM2.5 mapped with the GLA scale"></iframe>

## 1. The bundled colour scales

| Name | Pollutant | Bands |
|------|-----------|-------|
| `who_no2` (default) | NO2 | WHO guideline bands |
| `stripes_no2` | NO2 | stripe-style |
| `stripes_pm25` | PM2.5 | stripe-style |
| `airstat_no2` | NO2 | AirStat |
| `gla_pm25` | PM2.5 | GLA |
| `lbw_no2`, `lbrut_no2`, `lbm_no2` | NO2 | borough-specific (Wandsworth, Richmond, Merton) |
| `lbm_aqap_no2`, `lbm_aqap_pm10` | NO2, PM10 | Merton Air Quality Action Plan (print) |
| `deltas` | any | year-on-year change |
| `schools` | — | categorical (school type) |

`load_colour_scale("who_no2")` shows any scale's definition. (Two further
project-specific scales, `nine_elms` and `nine_elms_consolidated`, also
ship with the package.)

## 2. Anatomy of a scale — make your own

A scale is one YAML file: thresholds, colours, labels. Copy a bundled one
(`inst/config/scales/` in the package) and edit:

```yaml
name: my_no2
title: "NO2, µg/m³"
pollutant: NO2
thresholds: [0, 10, 20, 40, .Inf]
colours: ["green", "yellow", "orange", "red", "white"]
labels: ["< 10: Good", "10-20: Fair", "20-40: Poor", "> 40: Bad", "No data"]
```

The legend is generated from this file — labels with a description after
the range (like "10: Good") gain a footnote symbol and appear in the
legend's key row.

Two optional keys change how the legend reads:

```yaml
shape: circle             # the symbol drawn for this layer
footnote_symbols: false   # drop the † ‡ § markers, keep the key pills
```

`footnote_symbols: false` is worth setting for print. The markers pair a
ramp label with its pill, but each pill already carries its band's
colour, and at print size the markers are the first thing to become
illegible.

## 3. The figures on the legend

Every annual map shows the network **mean** for the year on the legend's
own ramp — a disc at its position, with the figure beside it — and the
**highest** site as a diamond. Colour already tells you the band, so the
two are told apart by shape.

The mean is over a fixed panel: only sites reporting in *every* year
shown are counted, so the figure moves with the air and not with the
network. Its caption gives the count. The maximum is the worst site
actually reporting that year, so its count can change between steps —
both captions say which basis they use.

```yaml
indicator:
  show: true              # the mean (default true)
  show_max: true          # and the maximum (default true)
  label: null             # null builds "Network mean, N sites"
  placement: "title_row"  # or "under_title", or "right" of the ramp
```

Narrowing `display_times` can *raise* the number of sites counted, since
a shorter run of years admits sites that did not report throughout.

**Annual maps only.** On a monthly or hourly map the figures do not
appear: the thresholds behind them are annual-mean limits, and marking a
month's mean against an annual limit would be wrong.

## 4. Themes — one file for the whole look

A theme bundles every visual choice so a series of maps stays consistent:

```{r theme}
quickmap(
  "tubes_merton.csv",
  boroughs   = "Merton",
  theme_file = system.file("themes", "merton.yaml", # NEW: a bundled theme
                           package = "quickmap"),
  output_file = "merton_styled.html"
)
```

<iframe src="https://ngnrfsk.github.io/quickmap/articles/maps/styling-theme.html" width="100%" height="440" loading="lazy" style="border:1px solid #ddd; border-radius:6px;" title="The Merton borough theme"></iframe>

The main sections (all optional — anything omitted keeps its default;
there is also a `legend:` section and a free-form `palette:` of named
colours):

```yaml
banner:
  style: "strip"          # "strip" (white, brand rule) or "bar" (colour block)
  background: "#5F3E94"   # the brand colour
  title: "Merton Air Quality"
map:
  vignette: true
  base_tiles: null        # null = OpenStreetMap; or e.g. "CartoDB.Positron"
  boundary_labels: false
  symbol_labels: false    # labels off unless a map asks for them
  label_scale: 1          # marker-label size (see the Labels chapter)
  label_background: true  # the plate behind each label
controls:
  autoplay: false
  play_speed: 500         # omit to let the step count decide
wind:                     # wind-particle styling (see the Wind chapter)
  particle_density: 0.0033
  line_width: 1.5
```

Bundled themes: `merton`, `wandsworth`, `richmond`, `airstat`,
`high_contrast` (`show_borough_colours()` lists them with their
palettes).

## 5. Who wins: parameter, theme, or default

Explicit `quickmap()` arguments beat the theme; the theme beats the
built-in defaults. So a theme sets the house style, and any single map
can override one thing:

```{r precedence, purl = FALSE}
quickmap(
  "tubes_merton.csv",
  boroughs      = "Merton",
  theme_file    = "my_theme.yaml",         # house style...
  banner_colour = "#078141",               # ...this map's banner differs
  output_file   = "merton_styled.html"
)
```

## 6. The background tiles

`map: base_tiles:` in the theme takes any
[Leaflet provider name](https://leaflet-extras.github.io/leaflet-providers/preview/).
The default (OpenStreetMap) carries the vignette well; the pale
`"CartoDB.Positron"` suits dense data.

## See also

- [Get started](quickmap.html) steps 4 and 7 — vignette colour and
  `banner_colour` in the build-up.
- [Wind](wind.html) — the theme's `wind:` section.
- `?load_theme`, `?load_colour_scale`, `?show_borough_colours`.
