---
title: "Time and animation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Time and animation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(eval = FALSE)
```

Any layer with more than one time step gets the time slider — drag it,
step with ‹ ›, press play. The steps come from the data: year columns in
a CSV animate annually, a datetime column in a sensor export animates at
its own resolution (hourly, daily, monthly). Nothing to configure:

```{r animation}
library(quickmap)                          # load the quickmap library
quickmap(
  from_rdata("episode_london.RData", # hourly sensor export
             "pm25", name = "Sensors"),
  boroughs     = c("Wandsworth", "Richmond"),
  colour_scale = "stripes_pm25",
  autoplay     = TRUE                      # play on open
)                                          # pace follows the step count
```

<iframe src="maps/getstarted-episode.html" width="100%" height="480" loading="lazy" style="border:1px solid #ddd; border-radius:6px;" title="108 hourly steps on the time slider"></iframe>

## 1. Choose which steps to show

`display_times` subsets to any steps present in the data, written the
way the data writes them:

| Resolution | Format | Example |
|------------|--------|---------|
| Annual | `"YYYY"` | `display_times = 2020:2022` |
| Monthly | `"YYYY-MM"` | `"2024-01"` |
| Daily | `"YYYY-MM-DD"` | `"2024-01-15"` |
| Hourly | `"YYYY-MM-DD HH:MM"` | `"2024-01-15 10:00"` |

```{r subset}
quickmap(
  "tubes_wandsworth.csv",
  boroughs      = "Wandsworth",
  display_times = 2020:2022                # NEW: three of the eight years
)
```

One selected step means no slider — a still map (that's how the static
export chooses its frame: [Get started](quickmap.html) step 6).

## 2. Autoplay and pace

`autoplay = TRUE` starts the animation when the map opens —
right for a public page; leave it off for a map people study.

`play_speed` is milliseconds per step. **You rarely need to set it.**
Left alone, the pace follows how many steps there are:

| Steps | Default pace |
|---|---|
| 12 or fewer (a run of years) | 1200 ms |
| 13 to 60 (months) | 800 ms |
| more than 60 (days, hours) | 450 ms |

A seven-year map at 1200 ms leaves nearly a second still on each year —
long enough to read the label and scan the symbols. A 108-step episode
would be tedious at that pace, so it runs faster.

## 3. The speed button

Whatever pace you choose, the reader can change it. The card carries a
button showing a multiplier — press it to cycle, and playback changes at
once:

- **Short animations** (12 steps or fewer) offer 0.5×, 1×, 2×, 4×.
- **Longer ones** also offer 0.25× and 8×, because there is more to
  fast-forward through.

It opens at 1×, so `play_speed` is always the pace the reader first
sees. On a screen narrower than 480px the button is hidden and the
theme's pace stands.

The colour fade between steps scales with the interval — 40% of it, up
to a quarter-second — so the map does not smear at 8× or crawl at
0.25×. Nothing to configure.

## 4. How big can an animation get?

- **Up to 200 steps** per map: beyond that QuickMap warns and keeps the
  most recent 200.
- **Above 50 steps** (or a large estimated file) the map switches
  automatically to a compact storage format — symbols are recoloured in
  place per step instead of duplicating every layer. The 108-step episode
  above is ~1 MB; the same map stored naively would be ~3.5 MB.
- Scrubbing stays instant regardless of step count (the recolour takes
  1–3 milliseconds).

## See also

- [Get started](quickmap.html) Part 2 — the animation build-up.
- [Wind](wind.html) — a wind overlay that advances with the slider.
- [Your data](your-data.html) — how time columns are recognised.
- [Styling and themes](styling.html) — the mean and maximum figures
  on the legend, which move with the slider.
