---
title: "Wind"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Wind}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(eval = FALSE)
```

Where did the pollution come from? An animated wind-particle overlay
answers that at a glance: streaks flow with the wind, coloured by speed,
and advance with the time slider. One parameter adds it:

```{r net-wind}
library(quickmap)                          # load the quickmap library
heathrow <- from_worldmet(station = "037720-99999", # Heathrow station id
                          year = 2024)              # fetches from NOAA
quickmap(
  from_rdata("episode_london.RData", "pm25", name = "Sensors"),
  boroughs     = c("Wandsworth", "Richmond"),
  colour_scale = "stripes_pm25",
  autoplay     = TRUE,
  play_speed   = 500,
  wind         = heathrow                  # the wind overlay
)
```

<iframe src="maps/getstarted-wind.html" width="100%" height="480" loading="lazy" style="border:1px solid #ddd; border-radius:6px;" title="Wind particles over the episode animation"></iframe>

## 1. Where the wind data comes from

`from_worldmet()` downloads hourly weather-station records from NOAA's
worldwide archive (via the `worldmet` package — install it and be online
for the fetch). Find a station near your map:

```{r net-getmeta}
worldmet::getMeta(lat = 51.47, lon = -0.45) # stations near Heathrow,
                                            # with their ids
```

For a city-scale map, one representative station is enough — the overlay
shows the period-average wind for each time step.

## 2. Bring your own wind data

Any data frame with `date`, `ws` (wind speed) and `wd` (wind direction)
columns works — from your own met station, a model, or a file:

```{r own-wind, purl = FALSE}
my_wind <- data.frame(
  date = as.POSIXct("2024-01-15 12:00", tz = "UTC") + 3600 * 0:23,
  ws   = 5,                                # m/s
  wd   = 225                               # degrees (from the south-west)
)
quickmap(
  from_rdata("episode_london.RData", "pm25", name = "Sensors"),
  boroughs = c("Wandsworth", "Richmond"),
  wind     = my_wind
)
```

Wind hours are matched to the map's time steps automatically (period
means for monthly or annual maps); steps with no wind data show no
particles rather than stale ones.

## 3. Styling the particles

The theme's `wind:` section controls the look
([Styling and themes](styling.html)):

```yaml
wind:
  colour_ramp: ["#4575b4", "#74add1", "#8fc3a7",   # blue -> red with speed
                "#fdae61", "#f46d43", "#d73027"]
  particle_density: 0.0033
  line_width: 1.5
  velocity_scale: 0.01
```

The default ramp colours particles by wind speed, starting at a visible
mid-blue so calm periods still read. On very busy backgrounds a darker
single-colour ramp or a heavier `line_width` stands out more.

## 4. Limits worth knowing

- **Interactive maps only** — a particle animation has no meaning in a
  static JPG, so exports omit it.
- The overlay is a uniform city-scale field from one station;
  multi-station and gridded (ERA5) wind fields are on the post-1.0
  roadmap.
- A background browser tab may pause the particles until it is
  foregrounded (a browser behaviour, not a defect).

## See also

- [Get started](quickmap.html) step 11 — wind in the build-up.
- [Time and animation](time.html) — the slider the particles follow.
- `?from_worldmet` — the full contract.
