---
title: "Your data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Your data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(eval = FALSE)
```

QuickMap reads four kinds of input: diffusion tube spreadsheets (CSV),
context CSVs such as school lists, sensor-network exports (RData), and
data frames you build in R — including live downloads via the
[OpenAir](https://openair-project.github.io/book/) tools. QuickMap
recognises which is which from the columns, never from the file name.
This page gives the exact requirements for each, then worked downloads
for when you have no file at all.

## 1. Diffusion tube CSVs

```{r csv, purl = FALSE}
library(quickmap)                           # load the quickmap library
quickmap("~/my_data/tubes_wandsworth.csv")  # tubes -> a map
```

| Easting | Northing | 2017 | 2018 | ... | 2024 | Label |
|---------|----------|------|------|-----|------|-------|
| 528900  | 172431   | 33   | 31   | ... | 15.9 | 11b Elmbourne Road |
| 530129  | 177727   |      | 49   | ... | 22.9 | 1 Nine Elms |

*Minimum required: `Easting`, `Northing` (British National Grid — QuickMap
converts them) and at least one four-digit year column. Blank cells are
fine: the site sits that year out.*

- A `Label` column supplies site names for labels
  ([Labels](labels.html)).
- **Gotcha:** if a `Label` column exists, rows whose label cell is empty
  are dropped from the map entirely.
- Any other column is carried along silently.

## 2. Context CSVs (schools and other places)

```{r schools, purl = FALSE}
quickmap(
  list("~/my_data/tubes_merton.csv",
       "~/my_data/schools_merton.csv"),     # context layer on top
  boroughs = "Merton",
  output_file = "my_first_map.html"
)
```

| School | Easting | Northing | Level |
|--------|---------|----------|-------|
| Abbotsbury Primary School | 525622 | 167783 | Primary |
| All Saints' CofE Primary School | 526473 | 170427 | Primary |

*Minimum required: `Easting`, `Northing`, `School`. A `School` column is
what makes QuickMap treat the file as a context layer (outline symbols,
no time steps); its text becomes the site's label. `Level` colours
schools categorically with the `schools` colour scale, and names the
categories in the banner key — which is what lets
`from_csv(..., shorten_labels = TRUE)` drop "Primary School" from every
marker ([Labels](labels.html), section 5).*

## 3. Sensor-network exports (RData)

```{r rdata}
library(quickmap)
quickmap(
  from_rdata("sensors_london.RData",
             "no2",                          # the pollutant column to map
             name = "Breathe London"),
  boroughs = "Merton",
  output_file = "my_first_map.html"
)
```

An RData file must contain a data frame in the long, OpenAir-style
layout — one row per site per time step:

| siteCode | year | no2  | pm25 | lat     | lon      |
|----------|------|------|------|---------|----------|
| CLDP0001 | 2021 | 28.1 | 11.2 | 51.4102 | -0.1907  |
| CLDP0001 | 2022 | 25.4 | 10.6 | 51.4102 | -0.1907  |

*Minimum required: `siteCode`, `lat`, `lon` (WGS84), a time column
(`year`, `date` or `year_str`) and the pollutant column you name.*

- QuickMap finds a compatible data frame inside the file by its columns:
  standard object names are checked first, then the largest match — so
  any object name works, and `from_rdata(data_object_name = )` picks one
  explicitly.
- Datetime columns set the animation's resolution automatically: hourly
  data animates hourly, annual data annually
  ([Time and animation](time.html)).

## 4. Data frames you built yourself

Anything already in R — a model output, a spreadsheet you cleaned, an API
result — becomes a layer with `qm_layer()`:

```{r hand-built}
my_sites <- data.frame(
  code = c("A", "B", "C"),                 # a stable id per site (required)
  lat  = c(51.44, 51.45, 51.46),           # latitude, WGS84 (required)
  lon  = c(-0.17, -0.19, -0.21),           # longitude, WGS84 (required)
  no2  = c(18, 27, 41)                     # the values to map (required)
)
quickmap(qm_layer(my_sites, name = "My survey"))
```

The full column contract (time columns, labels, value inference):
[Layers](layers.html) step 5 and `?qm_layer`.

## 5. No file at all: fetch data with the OpenAir tools

QuickMap is the spatial companion to
[OpenAir](https://openair-project.github.io/book/), whose packages
download UK (and worldwide) monitoring data. Each download drops
straight into `quickmap()`.

### Reference monitors (AURN) — `openair::importUKAQ()`

```{r net-aurn}
library(openair)                           # the OpenAir data tools
kc1 <- importUKAQ(site = "kc1",            # North Kensington site code
                  year = 2023)             # one year of hourly data
quickmap(
  from_openair(kc1,
               pollutant = "no2",          # pollutant column to map
               avg.time = "month",         # aggregate hours -> monthly means
               source = "aurn",            # network, for coordinate lookup
               name = "AURN"),
  boroughs = "All",
  output_file = "my_first_map.html"
)
```

Site codes and what each network measures:
[the OpenAir book](https://openair-project.github.io/book/sections/data-access/UK-air-quality-data.html).

### London networks (LAQN / Breathe London) — `openair::importImperial()`

```{r net-imperial, purl = FALSE}
laqn <- importImperial(site = "wm0",       # a LAQN site code
                       year = 2023)
quickmap(
  from_openair(laqn, pollutant = "no2", avg.time = "month",
               source = "kcl", name = "LAQN"),
  boroughs = "All",
  output_file = "my_first_map.html"
)
```

### Weather for the wind overlay — `worldmet::importNOAA()`

```{r net-noaa}
heathrow <- from_worldmet(station = "037720-99999", # Heathrow station id
                          year = 2024)              # fetches from NOAA
```

`from_worldmet()` wraps the download and conversion in one call; find
station ids with `worldmet::getMeta()`. Used in
[Get started](quickmap.html) step 11 and the [Wind](wind.html) chapter.

## 6. How recognition works

Content decides, not filenames (duck typing):

| The file has… | QuickMap treats it as |
|---|---|
| year columns (`2017`, `2018`, …) | temporal tube layer (solid symbols, time slider) |
| a `School` column | context layer (outline symbols, no time steps) |
| `siteCode` + pollutant + coordinates | sensor-network layer |

Rename your files freely; nothing is inferred from names. To overrule the
detection, build the layer explicitly with `from_csv(temporal = )` or
`qm_layer()`.

## See also

- [Get started](quickmap.html) — the first map, step by step.
- [Layers](layers.html) — combining these inputs on one map.
- `?from_csv`, `?from_rdata`, `?from_openair`, `?from_worldmet` — the
  full contracts.
